Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-05-02 paper.tex
Some updates to the James approach, and a simplification of the method.

The most natural extension of the monopole prescription is to include higher-order multipole moments. If the entire mass distribution is enclosed, then the potential can be expanded in a series of spherical harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the coordinate system is taken to be the center of the computational domain, and $r$ is the distance to the origin. The multipole moments can be calculated by expanding the Green's function for the Poisson equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation \ref{spherical_harmonic_expansion} using the addition theorem for spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty} \left[Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} + \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right].\label{multipole_potential}
\end{equation}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. The modified multipole moments are:
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
This approach becomes computationally feasible when we cut off the outer summation in Equation \ref{multipole_potential} at some finite value of $l_{\text{max}}$. If it is of sufficiently high order, we will accurately capture the distribution of mass on the grid. In practice we first evaluate the discretized analog of the modified multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m \leq l$, an operation that scales as $N^3$. We then directly compute the value of the potential on all of the $6N^2$ boundary zones. Since the multipole moments only need to be calculated once, the full operation scales only as $N^3$. The amount of time required to calculate the boundary conditions will be directly related to the chosen value of $l_{\text{max}}$, so there is a trade-off between computational expense and accuracy of the result. We find that for the early stages of the evolution of the binary system, $l_\text{max} \approx 6$ is sufficient.

@@

%Figure~\ref{fig:2levgrav} shows the sequence of updates for $\phi$ on a 2-level grid.
A clever way to evaluate Equation \ref{direct_sum} for the boundary values of the potential was described by \cite{james77}. In electrostatics a commonly encountered boundary-value problem is that of a positive charge distribution inside of a grounded conducting box. The potential vanishes on the surface of the box. For this to happen, a collection of negative charges must form on the outer surface of the box, to compensate the charge enclosed inside the box. This can be described by a surface charge density $\sigma$. This construction is useful in determining what the potential would be at the location of the surface when the box is removed. We will describe the potential outside of a density distribution that is non-zero only in some finite region as ``isolated'', and denote it by $\Phi$. This is the quantity we want to solve for. The potential in the case where the surface is grounded and conducting is called the ``homogeneous'' potential, and will be denoted by $\phi$. The homogeneous potential is the sum of the isolated potential and also the potential due to the surface charges $\Psi$:
\begin{equation*}
  \phi = \Phi + \Psi.
\end{equation*}
In order for $\phi$ to vanish at the surface, $\Phi$ and $\Psi$ must be of opposite sign, which makes sense since they are generated by opposite-sign charges. Rearranging this equation for the isolated potential yields
\begin{equation}
  \Phi = \phi - \Psi.
\end{equation}
Therefore a strategy for finding $\Phi$ is to solve the Poisson equation in the case where the boundary conditions are homogeneous and obtain $\phi$, and subtract from it the $\Psi$ that would be generated by the necessary compensating surface charges $\sigma$. All of this can be done for the gravitational version of Poisson's equation. Even though we do not commonly speak of negative gravitational charges, they are simply a mathematical tool for obtaining the correct potential. In general,
\begin{equation}
  \sigma = \frac{1}{4\pi G} {\bm{\nabla}} \phi \cdot \hat{\mathbf{n}},
\end{equation}
where $\hat{\mathbf{n}}$ is the unit vector perpendicular to (and pointing away from) the surface. On the computational domain, using an appropriate discretized representation of the ${\bm{\nabla}}$ operator, we calculate $\sigma$ at every boundary location. We can then find the resulting potential by summing over all of the surface charges:
\begin{equation}
  \Psi_{\text{l,m,n}} = \sum_{\text{l$^\prime$,m$^\prime$,n$^\prime$}} \frac{G\sigma_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}}{\left|\mathbf{x}_{\text{l,m,n}} - \mathbf{x}_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}\right|} dA_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}
\end{equation}

%In general,
%\begin{equation}
%  \sigma = 4\pi G

>>>>

The most natural extension of the monopole prescription is to include higher-order multipole moments. If the entire mass distribution is enclosed, then the potential can be expanded in a series of spherical harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the coordinate system is taken to be the center of the computational domain, and $r$ is the distance to the origin. The multipole moments can be calculated by expanding the Green's function for the Poisson equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation \ref{spherical_harmonic_expansion} using the addition theorem for spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty} \left[Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} + \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right].\label{multipole_potential}
\end{equation}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the outer summation in Equation \ref{multipole_potential} at some finite value of $l_{\text{max}}$. If it is of sufficiently high order, we will accurately capture the distribution of mass on the grid. In practice we first evaluate the discretized analog of the modified multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m \leq l$, an operation that scales as $N^3$. We then directly compute the value of the potential on all of the $6N^2$ boundary zones. Since the multipole moments only need to be calculated once, the full operation scales only as $N^3$. The amount of time required to calculate the boundary conditions will be directly related to the chosen value of $l_{\text{max}}$, so there is a trade-off between computational expense and accuracy of the result. We find that for the early stages of the evolution of the binary system, $l_\text{max} \approx 6$ is sufficient.

@@

%Figure~\ref{fig:2levgrav} shows the sequence of updates for $\phi$ on a 2-level grid.
A clever way to evaluate Equation \ref{direct_sum} exactly for the boundary values of the potential was described by \cite{james77}, and we have implemented a method inspired by that work here. In electrostatics a commonly encountered boundary-value problem is that of a positive charge distribution inside of a grounded conducting box. The potential vanishes on the surface of the box. For this to happen, a collection of negative charges must form on the outer surface of the box, to compensate the charge enclosed inside the box. This can be described by a surface charge density $\sigma$. This construction is useful in determining what the potential would be at the location of the surface when the box is removed. We will describe the potential outside of a density distribution that is non-zero only in some finite region as ``isolated'', and denote it by $\Phi$. This is the quantity we want to solve for. The potential in the case where the surface is grounded and conducting is called the ``homogeneous'' potential, and will be denoted by $\phi$. The homogeneous potential is the sum of the isolated potential and also the potential due to the surface charges $\Psi$:
\begin{equation*}
  \phi = \Phi + \Psi.
\end{equation*}
In order for $\phi$ to vanish at the surface, $\Phi$ and $\Psi$ must be of opposite sign, which makes sense since they are generated by opposite-sign charges. Rearranging this equation for the isolated potential yields
\begin{equation}
  \Phi = \phi - \Psi.
\end{equation}
Therefore a strategy for finding $\Phi$ is: solve the Poisson equation with homogeneous boundary conditions for $\phi$; calculate the necessary surface charges $\sigma$; find the resulting $\Psi$ on the boundary of the domain; solve the Laplace equation with inhomogeneous boundary conditions for $\Psi$ inside the domain; and then take $\Phi = \phi - \Psi$.

We slightly simplify this approach by observing that, since $\phi$ is by construction zero on the boundary, the value of $\Psi$ everywhere on the boundary must be equal in magnitude and opposite in sign to $\Phi$. Therefore the problem of finding the boundary values for $\Phi$ is reduced to solving the Poisson equation for $\phi$, constructing the surface charges $\sigma$, and then performing the appropriate integration of $-\sigma$ to get the boundary values. We then proceed with the Poisson solve as normal.

All of this can be done for the gravitational version of Poisson's equation. Even though we do not commonly speak of negative gravitational charges, they are simply a mathematical tool for obtaining the correct potential. In general,
\begin{equation}
  \sigma = \frac{1}{4\pi G} {\bm{\nabla}} \phi \cdot \hat{\mathbf{n}},
\end{equation}
where $\hat{\mathbf{n}}$ is the unit vector perpendicular to (and pointing away from) the surface. On the computational domain, using an appropriate discretized representation of the ${\bm{\nabla}}$ operator\footnote{The gradient of $\phi$ is discontinuous across the surface; it is zero just outside the surface, and non-zero just inside the surface. Therefore instead of using a difference centered on the boundary edge, we take a one-sided difference using the grid points just inside the domain. In order for the discretization to remain second-order accurate, data from the two closest grid points are used.}, we calculate $\sigma$ at every boundary location. We can then find the resulting potential on the boundary by summing over all of the surface charges:
\begin{equation}
  \Psi_{\text{l,m,n}} = \sum_{\text{l$^\prime$,m$^\prime$,n$^\prime$}} \frac{G\sigma_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}}{\left|\mathbf{x}_{\text{l,m,n}} - \mathbf{x}_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}\right|} dA_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}
\end{equation}
Here (l, m, n) refer only to boundary locations. The self-contribution from a zone must be excluded in this sum, since the denominator would diverge there.

As described so far, the method is not especially attractive: calculation of the boundary values requires an additional Poisson solve and a sum that scales as $N^4$. Furthermore, the latter sum faces problems with balancing of the computational work: only processors that have data on the boundary contribute to the summation, while the others remain idle. This does not compare well with the $N^3$ scaling of the multipole approach.
